---
title: "Introduction to GeneExpressionAnalysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to GeneExpressionAnalysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Welcome to GeneExpressionAnalysis

This package provides a pipeline for exploring, normalizing, clustering, and enriching gene expression data from `ExpressionSet` objects.

### 1. Reading Data

Start by loading the library and an example dataset. We will use the `airway` package dataset for this demonstration.

```{r setup, message=FALSE, warning=FALSE}
library(GeneExpressionAnalysis)
library(airway)

# Load the airway example data and convert to ExpressionSet
data(airway)
expression_set <- loadfromSumE(airway)
```

### 2. Exploratory Data Analysis (EDA) & Outliers

Before normalization, it is essential to inspect the data distribution, summary statistics, and identify potential outliers.

```{r eda, fig.width = 8, fig.height = 8}
# View expression summary statistics
expression_summary(expression_set)

# View raw data distribution
plot_boxplot(expression_set, log_scale = TRUE, title = "Raw Data Distribution")

# Detect outliers (genes and samples)
gene_outliers <- gene_outlier_detection(expression_set, threshold = 3)
sample_outliers <- sample_outlier_detection(expression_set, threshold = 3)

# Remove outliers (this handles both gene and sample detection internally, samples are not removed like genes but flagged)
exp_clean <- remove_outliers(expression_set, threshold = 3)
```

### 3. Normalization

Normalization ensures that expression levels are comparable across samples. This involves filtering low-expression genes, log transformation, and quantile normalization.

```{r normalization, fig.width = 8, fig.height = 8}
# Filter out low expression genes
exp_filtered <- filter_low_exp(exp_clean, min_count = 10)

# Log transform the data
exp_log <- log_transform(exp_filtered)

# Apply quantile normalization
exp_norm <- quantile_norm(exp_log)

# View density distribution across samples after normalization
plot_distr(exp_norm, title = "Normalized Data Distribution (Density)")

# View normalized distribution, which should be symmetrical and centered
plot_boxplot(exp_norm, title = "Normalized Data Distribution")
```

### 4. Clustering

Two clustering methods are provided: k-means and hierarchical clustering. Additionally, the optimal number of clusters for k-means can be determined using an elbow plot.

```{r clustering, message=FALSE, warning=FALSE, fig.width = 8, fig.height = 8}
# Plot the elbow to find optimal k
optimal_k(exp_norm, max_k = 10)

# Perform k-means clustering with k=5
km <- kmeans_clust(exp_norm, k_clusters = 5)

# Perform hierarchical clustering on top 50 most variable genes
gene_var <- apply(Biobase::exprs(exp_norm), 1, var)
exp_top <- exp_norm[order(gene_var, decreasing = TRUE)[1:50], ]
hc <- hierarchical_clust(exp_top)
```

### 5. Visualization

The results are visualized using PCA, heatmaps, and dendrograms.

```{r visualization, fig.width = 12, fig.height = 8}
# 1. PCA Plot
plot_PCA(exp_norm)

# 2. Heatmap of top variable genes
plot_heatmap(exp_norm, top_n_genes = 50)

# 3. Gene Dendrogram (from hierarchical clustering)
plot_den(hc$hclust)

# 4. Sample Dendrogram (all genes)
plot_sample_den(exp_norm)
```

### 6. Enrichment Analysis

The package supports GO Enrichment, KEGG Enrichment, and Gene Set Enrichment Analysis (GSEA) to look into the underlying biology of identified gene clusters.

```{r enrichment, message=FALSE, warning=FALSE, fig.width=8, fig.height=8, cache=TRUE}
cluster1_genes <- names(km$cluster[km$cluster == 1])

# 1. GO Enrichment
go_res <- go_enrich(cluster1_genes)
plot_enrichment(go_res, top_n = 10)

# 2. KEGG Enrichment
kegg_res <- kegg_enrich(cluster1_genes)
plot_enrichment(kegg_res, top_n = 10)

# 3. GSEA
gsea_res <- go_gse(exp_norm, condition_col = "dex", reference_level = "untrt")
plot_enrichment(gsea_res, top_n = 10)
```

### The All-In-One Pipeline

Instead of running the above steps manually, `GeneExpressionAnalysis` provides the `run_full_pipeline()` function to execute the entire analysis in a single command.

```{r full-pipeline, eval=FALSE}
results <- run_full_pipeline(
  expression_set = expression_set,
  k_clusters = 5,
  enrich_cluster = 1,
  condition_col = "dex",
  reference_level = "untrt"
)
```

### Session Info

```{r session-info}
sessionInfo()
```
